---
title: "Permutation test — cis-gene analysis (t-test + ash)"
author: "Yifan Zhou"
date: "6/27/2019"
output:  
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=3, fig.path='Figs/',
  echo=TRUE, warning=FALSE, message=FALSE, comment = NA, fig.align = 'center')
```

```{r load data}
library(ashr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(gridExtra)
options(stringsAsFactors = F)
wkdir <- '~/Downloads/ASoC/singlecell/'
load(paste0(wkdir,'data/cropseq_expression.Rd'))
vst_out <- readRDS(paste0(wkdir,'data/sctransformed_object.rds'))
sctransform_data <- vst_out$y
edgeR.res1Mb <- read.delim(paste0(wkdir,'data/edgeR_res.1Mb.txt'),header = T,sep = '\t')
edgeR.res500kb <- read.delim(paste0(wkdir,'data/edgeR_res.500kb.txt'),header = T,sep = '\t')
edgeR.res200kb <- read.delim(paste0(wkdir,'data/edgeR_res.200kb.txt'),header = T,sep = '\t')
```

## Categorical regression on permutated `sctransform`ed data

We still use the `sctransform`ed Pearson residuals, we filter out cells without any detected gRNA, and we focus only on the genes within $\pm$ 1 MB of the SNP loci of interest. This gives us a gene $\times$ cell matrix with the following dimensions:

```{r filter}
uniq.cisgenes.1Mb <- edgeR.res1Mb[!duplicated.default(edgeR.res1Mb$cisGene),c(2,4)]
subset.genes <- uniq.cisgenes.1Mb$cisGene
nonzero.cells <- colnames(exp.per.gRNA)[colSums(exp.per.gRNA>0)>0]
subset.sctransform_data <- sctransform_data[subset.genes,nonzero.cells]
cat(dim(subset.sctransform_data))
```

In the original gRNA-based DE analysis, for each gRNA target (70 in total), we partitioned the cells into 2 groups: cells that contain the target gRNA , and cells that don't contain the target gRNA.

In the permutation tests here, for each gRNA target, we divide the cells into 2 groups **randomly**, while keeping the number of cells in each group the same as before. Then, we conducted categorical regression on the Pearson residuals (based on the permuted conditions) to obtain the summary statistics. As before, 70 partitions were tested.

```{r categoric function}
categoric <- function(mtx,tg.cells,neg.cells){
  pval <- rep(NA,nrow(mtx))
  beta <- rep(NA,nrow(mtx))
  se <- rep(NA,nrow(mtx))
  subset <- mtx[,c(tg.cells,neg.cells)]
  for (i in 1:nrow(subset)){
    df <- data.frame(row.names = c(tg.cells,neg.cells),
                     residual = as.numeric(subset[i,]),
                     condition=c(rep(1,length(tg.cells)),
                                 rep(0,length(neg.cells))))
    cat.lm <- summary(lm(residual~factor(condition), data = df))
    beta[i] <- cat.lm$coefficients[2,1]
    se[i] <- cat.lm$coefficients[2,2]
    pval[i] <- cat.lm$coefficients[2,4]
  }
  names(beta) <- rownames(mtx)
  names(se) <- rownames(mtx)
  names(pval) <- rownames(mtx)
  return(list(beta=beta,se=se,pval=pval))
}
```

```{r permute function}
# test for each gRNA target (1 vs others, conditions permuted)
permute_categ <- function(seed){
  set.seed(seed)
  perm.summary <- list()
  for (glocus in row.names(exp.per.gRNA)[-c(51,72:76)]){
    # print(paste('gRNA target:',glocus))
    tg.cells <- colnames(exp.per.gRNA)[exp.per.gRNA[glocus,]>0]
    tg.indx <- sample(length(nonzero.cells),length(tg.cells))
    tg.cells <- nonzero.cells[tg.indx]
    neg.cells <- nonzero.cells[-tg.indx]
    perm.summary[[glocus]] <- categoric(subset.sctransform_data,tg.cells,neg.cells)
  }
  return(perm.summary)
}
```

```{r ash_on_ttest function}
ash_on_scttest <- function(uniq.cisgenes,option){
  for (i in 1:nrow(uniq.cisgenes)) {
    if (option=='enhancer'){
      tmp.column <- uniq.cisgenes$locus[i]
    } else if (option=='gRNA') {
      tmp.column <- uniq.cisgenes$gRNA[i]
    } else {
      print('Please provide a valid type: enhancer or gRNA.')
      return(1)
    }
    tmp.gene <- uniq.cisgenes$cisGene[i]
    if (is.na(match(tmp.gene,subset.genes))){ 
      next 
    } else {
      uniq.cisgenes$ttest.beta[i] <- beta.mtx[tmp.gene,tmp.column]
      uniq.cisgenes$ttest.se[i] <- se.mtx[tmp.gene,tmp.column]
      uniq.cisgenes$ttest.pval[i] <- pval.mtx[tmp.gene,tmp.column]
    }
  }
  uniq.cisgenes <- na.omit(uniq.cisgenes)
  cat(nrow(uniq.cisgenes),'data points:')
  uniq.cisgenes$zscore <- uniq.cisgenes$ttest.beta/uniq.cisgenes$ttest.se
  plot1 <- ggplot(uniq.cisgenes,aes(zscore)) + geom_histogram(aes(y=..density..),bins = 20) +
    labs(x = 'z-score from ttest',y='frequency') + 
    stat_function(fun = dnorm, args = list(mean = 0, sd = 1),col=2)
  
  ash.ttest <- ash(uniq.cisgenes$ttest.beta,uniq.cisgenes$ttest.se,mixcompdist = '-uniform')
  uniq.cisgenes$ash.beta <- ash.ttest$result$PosteriorMean
  uniq.cisgenes$lfsr <- get_lfsr(ash.ttest)
  log10_lfsr <- log10(get_lfsr(ash.ttest))
  plot2 <- qplot(uniq.cisgenes$ttest.beta, uniq.cisgenes$ash.beta,
                 color=log10_lfsr, main = 'ash shrinkage on ttest result',
                 xlab = 'log fold change', ylab = 'beta after shrinkage') + 
    geom_abline(intercept = 0,slope = 1,linetype = "dotted")
  grid.arrange(plot1,plot2,ncol=2)
  uniq.cisgenes <- uniq.cisgenes[order(uniq.cisgenes$lfsr),]
  uniq.cisgenes <- format(uniq.cisgenes,digits=3)
  print(kable(uniq.cisgenes,row.names = F) %>% kable_styling() %>%
          scroll_box(width = "100%", height = "200px"))
  ash.stats <- data.frame(pi=ash.ttest$fitted_g$pi, 
                        interval_a=ash.ttest$fitted_g$a, interval_b=ash.ttest$fitted_g$b)
  cat('ash estimated mixture proportions and corresponding intervals:')
  print(kable(ash.stats[ash.stats$pi!=0,], row.names = F, digits = 3) %>% 
          kable_styling(position = 'center',full_width = T) )
}
```

## Permutation 1

```{r perm1}
summary_per_gene <- permute_categ(1952) # five random seeds: 1952 8076  966  275 7816
saveRDS(summary_per_gene,file = paste0(wkdir,'data/cis.categ.perm.summary01.rds'))
```

```{r get summary1}
beta.mtx <- sapply(summary_per_gene, function(x){x$beta})
se.mtx <- sapply(summary_per_gene, function(x){x$se})
pval.mtx <- sapply(summary_per_gene, function(x){x$pval})
```

### ash on 200 kb _cis_ genes
```{r perm1.200kb, results='asis'}
cisgenes.200kb <- edgeR.res200kb[,c(1:4)]
ash_on_scttest(cisgenes.200kb,'gRNA')
```

***

### ash on 500 kb _cis_ genes
```{r perm1,500kb, results='asis'}
cisgenes.500kb <- edgeR.res500kb[,c(1:4)]
ash_on_scttest(cisgenes.500kb,'gRNA')
```

***

### ash on 1 Mb _cis_ genes
```{r perm1.1Mb, results='asis'}
cisgenes.1Mb <- edgeR.res1Mb[,c(1:4)]
ash_on_scttest(cisgenes.1Mb,'gRNA')
```

***

## Permutation 2

```{r perm2}
summary_per_gene <- permute_categ(8076) # five random seeds: 1952 8076  966  275 7816
saveRDS(summary_per_gene,file = paste0(wkdir,'data/cis.categ.perm.summary02.rds'))
beta.mtx <- sapply(summary_per_gene, function(x){x$beta})
se.mtx <- sapply(summary_per_gene, function(x){x$se})
pval.mtx <- sapply(summary_per_gene, function(x){x$pval})
```

### ash on 200 kb _cis_ genes
```{r perm2.200kb, results='asis'}
cisgenes.200kb <- edgeR.res200kb[,c(1:4)]
ash_on_scttest(cisgenes.200kb,'gRNA')
```

***

### ash on 500 kb _cis_ genes
```{r perm2.500kb, results='asis'}
cisgenes.500kb <- edgeR.res500kb[,c(1:4)]
ash_on_scttest(cisgenes.500kb,'gRNA')
```

***

### ash on 1 Mb _cis_ genes
```{r perm2.1Mb, results='asis'}
cisgenes.1Mb <- edgeR.res1Mb[,c(1:4)]
ash_on_scttest(cisgenes.1Mb,'gRNA')
```

***

## Permutation 3

```{r perm3}
summary_per_gene <- permute_categ(966) # five random seeds: 1952 8076  966  275 7816
saveRDS(summary_per_gene,file = paste0(wkdir,'data/cis.categ.perm.summary03.rds'))
beta.mtx <- sapply(summary_per_gene, function(x){x$beta})
se.mtx <- sapply(summary_per_gene, function(x){x$se})
pval.mtx <- sapply(summary_per_gene, function(x){x$pval})
```

### ash on 200 kb _cis_ genes
```{r perm3.200kb, results='asis'}
cisgenes.200kb <- edgeR.res200kb[,c(1:4)]
ash_on_scttest(cisgenes.200kb,'gRNA')
```

***

### ash on 500 kb _cis_ genes
```{r perm3.500kb, results='asis'}
cisgenes.500kb <- edgeR.res500kb[,c(1:4)]
ash_on_scttest(cisgenes.500kb,'gRNA')
```

***

### ash on 1 Mb _cis_ genes
```{r perm3.1Mb, results='asis'}
cisgenes.1Mb <- edgeR.res1Mb[,c(1:4)]
ash_on_scttest(cisgenes.1Mb,'gRNA')
```

## Permutation 4

```{r perm4}
summary_per_gene <- permute_categ(275) # five random seeds: 1952 8076  966  275 7816
saveRDS(summary_per_gene,file = paste0(wkdir,'data/cis.categ.perm.summary04.rds'))
beta.mtx <- sapply(summary_per_gene, function(x){x$beta})
se.mtx <- sapply(summary_per_gene, function(x){x$se})
pval.mtx <- sapply(summary_per_gene, function(x){x$pval})
```

### ash on 200 kb _cis_ genes
```{r perm4.200kb, results='asis'}
cisgenes.200kb <- edgeR.res200kb[,c(1:4)]
ash_on_scttest(cisgenes.200kb,'gRNA')
```

***

### ash on 500 kb _cis_ genes
```{r perm4.500kb, results='asis'}
cisgenes.500kb <- edgeR.res500kb[,c(1:4)]
ash_on_scttest(cisgenes.500kb,'gRNA')
```

***

### ash on 1 Mb _cis_ genes
```{r perm4.1Mb, results='asis'}
cisgenes.1Mb <- edgeR.res1Mb[,c(1:4)]
ash_on_scttest(cisgenes.1Mb,'gRNA')
```

## Permutation 5

```{r perm5}
summary_per_gene <- permute_categ(7816) # five random seeds: 1952 8076  966  275 7816
saveRDS(summary_per_gene,file = paste0(wkdir,'data/cis.categ.perm.summary05.rds'))
beta.mtx <- sapply(summary_per_gene, function(x){x$beta})
se.mtx <- sapply(summary_per_gene, function(x){x$se})
pval.mtx <- sapply(summary_per_gene, function(x){x$pval})
```

### ash on 200 kb _cis_ genes
```{r perm5.200kb, results='asis'}
cisgenes.200kb <- edgeR.res200kb[,c(1:4)]
ash_on_scttest(cisgenes.200kb,'gRNA')
```

***

### ash on 500 kb _cis_ genes
```{r perm5.500kb, results='asis'}
cisgenes.500kb <- edgeR.res500kb[,c(1:4)]
ash_on_scttest(cisgenes.500kb,'gRNA')
```

***

### ash on 1 Mb _cis_ genes
```{r perm5.1Mb, results='asis'}
cisgenes.1Mb <- edgeR.res1Mb[,c(1:4)]
ash_on_scttest(cisgenes.1Mb,'gRNA')
```
