---
title: "Permutation test -- trans-gene analysis (t-test + ash)"
author: "Yifan Zhou"
date: "6/16/2019"
output:   
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, echo=TRUE, warning=FALSE, 
                      message=FALSE, comment = NA, fig.align = 'center')
```

```{r load data}
library(ashr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(gridExtra)
options(stringsAsFactors = F)
wkdir <- '~/Downloads/ASoC/singlecell/'
load(paste0(wkdir,'data/cropseq_expression.Rd'))
vst_out <- readRDS(paste0(wkdir,'data/sctransformed_object.rds'))
sctransform_data <- vst_out$y
nlocus <- colSums(exp.per.enhancer>0)
gene.percent <- rowMeans(gene.exp>0)
ncell_uniq <- rep(NA,nrow(exp.per.gRNA))
for (i in 1:nrow(exp.per.gRNA)){
  glocus <- row.names(exp.per.gRNA)[i]
  ncell_uniq[i]=sum(exp.per.gRNA[glocus,]>0 & nlocus==1)
}
tmp <- sapply(strsplit(row.names(exp.per.gRNA),split = '_'), 
              function(x){paste(x[1],x[2],sep = '_')})
names(ncell_uniq) <- tmp
```

## DE analysis on permutated `sctransform`ed data

We still use the `sctransform`ed Pearson residuals, and we filtered the genes to be those present in at least 20% of all cells (8117 in total). 

For each target gRNA, we selected 2 groups of cells: cells that only contain the target gRNAs, and cells that only contain the negative control gRNAs. Then, we **permute the group labels** and conducted categorical regression of the Pearson residuals (against the 2 permuted conditions) to obtain the summary statistics. As before, 70 (gRNA vs neg ctrl) pairs were tested.

```{r categoric, eval=F}
categoric <- function(mtx,tg.cells,neg.cells){
  pval <- rep(NA,nrow(mtx))
  beta <- rep(NA,nrow(mtx))
  se <- rep(NA,nrow(mtx))
  subset <- mtx[,c(tg.cells,neg.cells)]
  for (i in 1:nrow(subset)){
    df <- data.frame(row.names = c(tg.cells,neg.cells),
                     residual = as.numeric(subset[i,]),
                     condition=c(rep(1,length(tg.cells)),
                                 rep(0,length(neg.cells))))
    cat.lm <- summary(lm(residual~factor(condition), data = df))
    beta[i] <- cat.lm$coefficients[2,1]
    se[i] <- cat.lm$coefficients[2,2]
    pval[i] <- cat.lm$coefficients[2,4]
  }
  return(list(beta=beta,se=se,pval=pval))
}
## Filter the genes to be present in > 20% cells:
subset.genes <- names(gene.percent)[gene.percent>0.2]
subset.sctransform_data <- sctransform_data[subset.genes,]
dim(subset.sctransform_data)
## Categorical regression on sctransformed data 
# test btw each gRNA condition and neg ctrl (conditions permuted)
neg.cells <- colnames(exp.per.enhancer)[exp.per.enhancer["neg",]>0 & nlocus==1]
perm.summary <- list()
for (glocus in row.names(exp.per.gRNA)[-c(51,72:76)]){
  # glocus <- "BAG5_1_gene"
  # print(paste('Permutation test, gRNA target:',glocus))
  tg.cells <- colnames(exp.per.gRNA)[exp.per.gRNA[glocus,]>0 & nlocus==1]
  combined.cells <- c(tg.cells,neg.cells)
  combined.cells <- combined.cells[sample(length(combined.cells))]
  perm.summary[[glocus]] <- categoric(subset.sctransform_data, combined.cells[1:length(tg.cells)],
                                      combined.cells[(length(tg.cells)+1):length(combined.cells)])
}
saveRDS(perm.summary,file = paste0(wkdir,'data/categ.perm.summary01.rds'))
```

## ashr meta-analysis on regression summary statistics

For each gene, we combined the 70 DE effect sizes and standard errors estimated in the previous step, and used an adaptive shrinkage method, `ashr`, to discern which effect sizes are truly non-zero (significant). Here, we used a cutoff of 0.1 for `lfsr` (the local false sign rate).
```{r summarize, eval=F}
perm.summary <- readRDS(paste0(wkdir,'data/categ.perm.summary01.rds'))
beta.mtx <- sapply(perm.summary, function(x){x$beta})
se.mtx <- sapply(perm.summary, function(x){x$se})
pval.mtx <- sapply(perm.summary, function(x){x$pval})
subset.genes <- names(gene.percent)[gene.percent>0.2]
rownames(beta.mtx) <- subset.genes
rownames(se.mtx) <- subset.genes
rownames(pval.mtx) <- subset.genes
```

```{r ashr, eval=F}
lfsr.mtx <- data.frame(matrix(nrow = nrow(beta.mtx),ncol = ncol(beta.mtx)),
                       row.names = rownames(beta.mtx))
names(lfsr.mtx) <- colnames(beta.mtx)
betaest.mtx <- lfsr.mtx
for (i in 1:nrow(beta.mtx)){
  ash.tmp <- ash(beta.mtx[i,],se.mtx[i,])
  lfsr.mtx[i,] <- get_lfsr(ash.tmp)
  betaest.mtx[i,] <- get_pm(ash.tmp)
}
save(betaest.mtx,lfsr.mtx,file = paste0(wkdir,'data/perm.ashr_estimation01.Rdata'))
```

## Permutation 1

```{r}
load(paste0(wkdir,'data/perm.ashr_estimation01.Rdata'))
tmp.gRNA <- colSums(lfsr.mtx<0.1)
tmp.gene <- rowSums(lfsr.mtx<0.1)
signif_genes <- row.names(betaest.mtx)[tmp.gene>0]
cat('There are',length(signif_genes),'significant genes in total.')
```

### gRNA targets per significant gene

Here are all the significant genes ($lfsr<0.1$) and the gRNA groups where they are differentially expressed:
```{r gene-gRNA}
signif_genes.gRNA.lst <- list()
for(i in signif_genes){
  gRNA.group <- colnames(betaest.mtx)[lfsr.mtx[i,]<0.1]
  signif_genes.gRNA.lst[[i]] <- sapply(strsplit(gRNA.group,split = '_'), 
                                       function(x){paste(x[1],x[2],sep = '_')})
}
# Make table:
compact.genes.lst <- sapply(signif_genes.gRNA.lst,FUN = function(x){paste(x,collapse = ', ')})
compact.genes.df <- as.data.frame(compact.genes.lst)
tmp.indx <- order(rownames(compact.genes.df))
compact.genes.df <- data.frame(signif_gene=rownames(compact.genes.df)[tmp.indx],
                               gRNA_targets=compact.genes.df[tmp.indx,])
# Present table:
kable(compact.genes.df) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
num_gRNA_pergene <- sapply(signif_genes.gRNA.lst,FUN = function(x){length(x)})
qplot(num_gRNA_pergene,geom = 'histogram',xlab = 'Number of gRNA targets per gene',ylab = 'Count',
     main = paste('Distribution of', length(num_gRNA_pergene), 'significant genes'),bins=10)
```

### Significant genes per gRNA target
```{r gRNA-gene}
gRNA.signif_genes.lst <- list()
for (i in signif_genes){
  cells <- signif_genes.gRNA.lst[[i]]
  for (c in cells){
    if (is.na(match(c, names(gRNA.signif_genes.lst)))){
      gRNA.signif_genes.lst[[c]] <- i
    } else {
      gRNA.signif_genes.lst[[c]] <- c(gRNA.signif_genes.lst[[c]],i)
    }
  }
}
# Make table:
compact.gRNAs.lst <- sapply(gRNA.signif_genes.lst,FUN = function(x){paste(x,collapse = ', ')})
compact.gRNAs.df <- as.data.frame(compact.gRNAs.lst)
tmp.indx <- order(rownames(compact.gRNAs.df))
compact.gRNAs.df <- data.frame(gRNA_target=rownames(compact.gRNAs.df)[tmp.indx],
                               signif_genes=compact.gRNAs.df[tmp.indx,])
compact.gRNAs.df$gRNA_target <- sapply(strsplit(compact.gRNAs.df$gRNA_target,split = '_'), function(x){paste(x[1],x[2],sep = '_')})
compact.gRNAs.df$num_cells <- ncell_uniq[compact.gRNAs.df$gRNA_target]
compact.gRNAs.df <- compact.gRNAs.df[,c(3,1,2)]
# Present table:
kable(compact.gRNAs.df) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
num_gene_pergRNA <- sapply(gRNA.signif_genes.lst,FUN = function(x){length(x)})
qplot(num_gene_pergRNA,geom = 'histogram',xlab = 'Number of DE genes per gRNA target',ylab = 'Count',
     main = paste('Distribution of', length(num_gene_pergRNA), 'gRNAs'),bins=20)
```

gRNA conditions with high number of significant genes are the ones with few cells compared to the number of negative control cells (111).

```{r}
qplot(compact.gRNAs.df$num_cells,num_gene_pergRNA[compact.gRNAs.df$gRNA_target],
      xlab = 'Number of cells in the target condition',ylab = 'Number of significant genes')
```

```{r result func, echo=F}
get.result <- function(betaest.mtx,lfsr.mtx){
  
  tmp.gRNA <- colSums(lfsr.mtx<0.1)
  tmp.gene <- rowSums(lfsr.mtx<0.1)
  signif_genes <- row.names(betaest.mtx)[tmp.gene>0]
  cat('There are',length(signif_genes),'significant genes in total.')
  
  signif_genes.gRNA.lst <- list()
  for(i in signif_genes){
    gRNA.group <- colnames(betaest.mtx)[lfsr.mtx[i,]<0.1]
    signif_genes.gRNA.lst[[i]] <- sapply(strsplit(gRNA.group,split = '_'), 
                                         function(x){paste(x[1],x[2],sep = '_')})
  }
  # Make table:
  compact.genes.lst <- sapply(signif_genes.gRNA.lst,FUN = function(x){paste(x,collapse = ', ')})
  compact.genes.df <- as.data.frame(compact.genes.lst)
  tmp.indx <- order(rownames(compact.genes.df))
  compact.genes.df <- data.frame(signif_gene=rownames(compact.genes.df)[tmp.indx],
                                 gRNA_targets=compact.genes.df[tmp.indx,])
  # Present table:
  print(kable(compact.genes.df) %>% kable_styling() %>% scroll_box(width = "100%", height = "300px"))
  num_gRNA_pergene <- sapply(signif_genes.gRNA.lst,FUN = function(x){length(x)})
  # qplot(num_gRNA_pergene,geom = 'histogram',xlab = 'Number of gRNA targets per gene',ylab = 'Count',
  #       main = paste('Distribution of', length(num_gRNA_pergene), 'significant genes'),bins=10)
  
  ### Significant genes per gRNA target
  gRNA.signif_genes.lst <- list()
  for (i in signif_genes){
    cells <- signif_genes.gRNA.lst[[i]]
    for (c in cells){
      if (is.na(match(c, names(gRNA.signif_genes.lst)))){
        gRNA.signif_genes.lst[[c]] <- i
      } else {
        gRNA.signif_genes.lst[[c]] <- c(gRNA.signif_genes.lst[[c]],i)
      }
    }
  }
  # Make table:
  compact.gRNAs.lst <- sapply(gRNA.signif_genes.lst,FUN = function(x){paste(x,collapse = ', ')})
  compact.gRNAs.df <- as.data.frame(compact.gRNAs.lst)
  tmp.indx <- order(rownames(compact.gRNAs.df))
  compact.gRNAs.df <- data.frame(gRNA_target=rownames(compact.gRNAs.df)[tmp.indx],
                                 signif_genes=compact.gRNAs.df[tmp.indx,])
  compact.gRNAs.df$gRNA_target <- sapply(strsplit(compact.gRNAs.df$gRNA_target,split = '_'),
                                         function(x){paste(x[1],x[2],sep = '_')})
  compact.gRNAs.df$num_cells <- ncell_uniq[compact.gRNAs.df$gRNA_target]
  compact.gRNAs.df <- compact.gRNAs.df[,c(3,1,2)]
  # Present table:
  print(kable(compact.gRNAs.df) %>% kable_styling() %>% scroll_box(width = "100%", height = "300px"))
  num_gene_pergRNA <- sapply(gRNA.signif_genes.lst,FUN = function(x){length(x)})
  plot1 <- qplot(num_gRNA_pergene,geom = 'histogram',xlab = 'Number of gRNA targets per gene',ylab = 'Count',
        main = paste('Distribution of', length(num_gRNA_pergene), 'significant genes'),bins=10)
  plot2 <- qplot(num_gene_pergRNA,geom = 'histogram',xlab = 'Number of DE genes per gRNA target',
        ylab = 'Count', main = paste('Distribution of', length(num_gene_pergRNA), 'gRNAs'),bins=20)
  plot3 <- qplot(compact.gRNAs.df$num_cells,num_gene_pergRNA[compact.gRNAs.df$gRNA_target],
      xlab = 'Number of cells in the target condition',ylab = 'Number of significant genes')
  grid.arrange(plot1,plot2,plot3,ncol=2)
}
```

## Permutation 2
```{r perm2,results='asis',fig.height=6,fig.width=9}
load(paste0(wkdir,'data/perm.ashr_estimation02.Rdata'))
get.result(betaest.mtx,lfsr.mtx)
```

## Permutation 3
```{r perm3,results='asis',fig.height=6,fig.width=9}
load(paste0(wkdir,'data/perm.ashr_estimation03.Rdata'))
get.result(betaest.mtx,lfsr.mtx)
```

## Permutation 4
```{r perm4,results='asis',fig.height=6,fig.width=9}
load(paste0(wkdir,'data/perm.ashr_estimation04.Rdata'))
get.result(betaest.mtx,lfsr.mtx)
```

## Permutation 5
```{r perm5,results='asis',fig.height=6,fig.width=9}
load(paste0(wkdir,'data/perm.ashr_estimation05.Rdata'))
get.result(betaest.mtx,lfsr.mtx)
```

## Session Information
```{r session}
sessionInfo()
```
